Supermartingales and Termination Analysis of Statistical PCF

### 2020-12-26 LICS21/probabilisticProgrammingMartingales.tex
Created version for LICS21 submission in subdirectory LICS21

\input{Style/comment}
\usepackage{calculation}
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{positioning, graphs, quotes, graphdrawing, shapes.geometric}
\usegdlibrary{layered, circular}
\usepackage[square]{natbib}
\setcitestyle{aysep={}}

@@

\paragraph{}
This definition of almost-sure termination is equivalent to that given in \citep{MakOP20b}, although the program semantics is stated in a slightly different way. In particular, the argument to $\tscore$ is not relevant to termination (except for the possibility that its argument's evaluation wouldn't terminate).

\iffalse
\lo{Your operational semantics does not maintain a record of the current weight of the reduction.
A.s.~termination does depend on $\tscore$: see \cite[\S 4.3]{DBLP:journals/corr/abs-2004-03924}\footnote{\url{https://arxiv.org/abs/2004.03924}}.
I think it important to take the behaviour of $\tscore$ into account;
you should do it as a future task.}
\fi

\section{Supermartingales}
One approach to proving that a term terminates almost surely is to find some variant that is bounded below and, on average, decreases sufficiently quickly that it must eventually reach 0, similarly to the approach taken in \citep{DBLP:journals/pacmpl/McIverMKK18} for imperative programs.

@@

For $M_2$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \log(x+1) + 1$.
Using shorthand 
$\Theta_2 = \tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
we can define an antitone partial ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ 
%with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
defined as: for $n \in \mathbb N$
\begin{align*}
{\Theta_2} \, n 
&\mapsto 
g_2(n)
\\
0 &\mapsto 0.
\end{align*}
For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$, each with probability $1/2$, with one $\tY$-reduction.
Now
\begin{align*}
  & g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
  = & \log \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) \\
  = & \frac 1 2 \log\left(\frac{(n+1)^2}{n(n+2)}\right) \\
  = & \frac 1 2 \log\left(1 + \frac 1 {n(n+2)}\right) \\
  > & \frac 1 {4n(n+2)} \\
  > & \frac 1 2 \frac 1 {5(n+1)^2} + \frac 1 2 \frac 1 {5(n+3)^2}
\end{align*}
and $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
Therefore by setting $\epsilon_2(x) = \frac 1 {5(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met. It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone partial ranking function, and $M_2$ is AST.

@@

\begin{example}[Fair-in-the-limit random walk]
\label{ex:Fair-in-the-limit random walk}\citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}
\[
\underbrace{\big
(\tY \, f \, x . 
\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)}_{\Xi} 
\, 10
\]
To construct an antitone ranking function, we solve the recurrence relation:
\[
\left\{
\begin{array}{rll}
z_0 &=& 0\\
n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
\end{array}
\right.
\]
For $n > 2$, $z_n = \log(n-1)$ works. The expected decrease is $\frac 1 2(\log(1+\frac 1 {(n-1)^2-1}) - \frac 1 {2n+1} \log(1 + \frac 2 {n-2}))$, and using the fact that $\frac x {x+1} \leq \log (1+x) \leq x$ for $x > 0$, this is at least $\frac 1 2(\frac 1 {(n-1)^2} - \frac 1 {2n + 1} \frac 2 {n-2}) = \frac {n^2}{2(n-1)^2(2n+1)(n-2)}$, which (again for $n > 2$) is positive and antitone. For $n = 0, 1, 2$, we take $\epsilon$ to be 9/40 (the same as its value at 3), then set $z_2 = 2 \log 2 - \log 3 - 1, z_1 = 3 \log 2 - 2 \log 3 - 3, z_0 = 4 \log 2 - 3 \log 3 - 6$. Some of those values are negative, but it's still bounded below, so by adding a constant offset it can be corrected, and the term is AST.

@@

\bibliographystyle{apalike}
\bibliography{references}

>>>>

\input{../Style/comment}
\usepackage{calculation}
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{positioning, graphs, quotes, graphdrawing, shapes.geometric}
\usegdlibrary{layered, circular}
\usepackage[square]{natbib}
\setcitestyle{aysep={}}

@@

\paragraph{}
This definition of almost-sure termination is equivalent to that given in \citep{MakOPW21}, although the program semantics is stated in a slightly different way. In particular, the argument to $\tscore$ is not relevant to termination (except for the possibility that its argument's evaluation wouldn't terminate).

\iffalse
\lo{Your operational semantics does not maintain a record of the current weight of the reduction.
A.s.~termination does depend on $\tscore$: see \cite[\S 4.3]{MakOPW21}\footnote{\url{https://arxiv.org/abs/2004.03924}}.
I think it important to take the behaviour of $\tscore$ into account;
you should do it as a future task.}
\fi

\section{Supermartingales}
One approach to proving that a term terminates almost surely is to find some variant that is bounded below and, on average, decreases sufficiently quickly that it must eventually reach 0, similarly to the approach taken in 
%\citep{DBLP:journals/pacmpl/McIverMKK18} 
\cite{DBLP:conf/popl/FioritiH15} for imperative programs.

@@

For $M_2$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \log(x+1) + 1$.
Using shorthand 
$\Theta_2 = \tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
we can define an antitone partial ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ 
%with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
defined as: for $n \in \mathbb N$
\begin{align*}
{\Theta_2} \, n 
&\mapsto 
g_2(n)
\\
0 &\mapsto 0.
\end{align*}
For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$, each with probability $1/2$, with one $\tY$-reduction.
Now
\begin{align*}
g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 
  &=  \log \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) \\
  &=  \frac 1 2 \log\left(\frac{(n+1)^2}{n(n+2)}\right) \\
  &=  \frac 1 2 \log\left(1 + \frac 1 {n(n+2)}\right) \\
  &>  \frac 1 {4n(n+2)} \\
  &>  \frac 1 2 \frac 1 {5(n+1)^2} + \frac 1 2 \frac 1 {5(n+3)^2}
\end{align*}
\changed[lo]{(The first inequality follows from the fact that for all $x \in (0, 1)$, $\log (1 + x) > x$.)}
Moreover $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
Therefore by setting $\epsilon_2(x) = \frac 1 {5(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met. It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone partial ranking function, and $M_2$ is AST.

@@

\begin{example}[Fair-in-the-limit random walk]
\label{ex:Fair-in-the-limit random walk}\citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}
\[
\underbrace{\big
(\tY \, f \, x . 
\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)}_{\Xi} 
\, 10
\]
To construct an antitone ranking function, we solve the recurrence relation:
\[
\left\{
\begin{array}{rll}
z_0 &=& 0\\
n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
\end{array}
\right.
\]
For $n > 2$, $z_n = \log(n-1)$ works. The expected decrease is $\frac 1 2(\log(1+\frac 1 {(n-1)^2-1}) - \frac 1 {2n+1} \log(1 + \frac 2 {n-2}))$, and using the fact that $\frac x {x+1} \leq \log (1+x) \leq x$ for $x > 0$,\lo{The preceding inequality is incorrect: for $x \in (0, 1)$, $\log (1 + x) > x$.} this is at least $\frac 1 2(\frac 1 {(n-1)^2} - \frac 1 {2n + 1} \frac 2 {n-2}) = \frac {n^2}{2(n-1)^2(2n+1)(n-2)}$, which (again for $n > 2$) is positive and antitone. For $n = 0, 1, 2$, we take $\epsilon$ to be 9/40 (the same as its value at 3), then set $z_2 = 2 \log 2 - \log 3 - 1, z_1 = 3 \log 2 - 2 \log 3 - 3, z_0 = 4 \log 2 - 3 \log 3 - 6$. Some of those values are negative, but it's still bounded below, so by adding a constant offset it can be corrected, and the term is AST.

@@

\bibliographystyle{apalike}
\bibliography{../references}

### 2020-12-26 LICS21/ranking-supermartingales.tex
Created version for LICS21 submission in subdirectory LICS21

\begin{definition}\rm
A sequence of random variables $(Y_n)_{n \in \omega}$ adapted to a filtration $(\calF_n)_{n \in \omega}$ is a \emph{supermartingale} if for all $n \in \omega$, $\expect{|Y_n|} < \infty$, and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$.
\lo{The preceding means: $\forall A \in \calF_n$, $\int_A \dif \mu \, Y_{n+1} \leq \int_A \dif \mu \,Y_n$.}
For $\epsilon > 0$ and a stopping time $T$ w.r.t.~$(\calF_n)_{n \in \omega}$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
A \emph{ranking supermartingale} is a $\epsilon$-ranking supermartingale for some $\epsilon > 0$.
\end{definition}

\begin{remark}
\changed[lo]{The original definition of ranking supermartingale, due to \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, does not mention stopping time explicitly:
Fioriti and Hermanns call a supermartingale $(Y_n)_{n \in \omega}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot 1_{\set{Y_n > 0}}$.}
\end{remark}

@@

The following is an iterated version of Doob's well-known Optional Sampling Theorem (see, e.g., \cite[\S 6.7]{AshDD00}).
\begin{theorem}[Optional Sampling]
\label{thm:optional sampling}
Let $(X_n)_{n\in \omega}$ be a supermartingale, and $(T_n)_{n \in \omega}$ a sequence of increasing stopping times, then $(X_{T_n})_{n \in \omega}$ is a supermartingale adapted to $(\calF_{T_n})_{n \in \omega}$ if one of the following conditions holds:
\begin{enumerate}
\item each $T_n$ is bounded i.e.~$T_n < c_n$ where $c_n$ is a constant
\item $(X_n)_{n\in \omega}$ is uniformly integrable.
\end{enumerate}
\end{theorem}

\begin{lemma}[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}}]
% \citep[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}
\label{lem:rank-PAST}
Let $(Y_n)_{n \in \omega}$ be a $\epsilon$-ranking supermartingale with respect to the stopping time $T$, then $T < \infty$ almost surely, and $\expect{T} \leq \dfrac{\expect{Y_0}}{\epsilon}$.
\end{lemma}

\begin{proof}[Proof Sketch]
First establish by induction on $n$: 
\[
\forall n \in \omega \, . \, \expect{Y_n} \leq \expect{Y_0} - \epsilon \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[T > i]\Big).
\]
It follows that $\sum_{i=0}^{\infty} \mathbb{P}[T > i]$ converges; hence $\lim_{i \to \infty} \mathbb{P}[T > i] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
Then it suffices to prove: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=0} \mathbb{P}[T > i]$.
\end{proof}

\lo{\Cref{lem:rank-PAST} is a slight extension of \citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}. In op.~cit., the stopping time $T$ is assumed to be the random variable $s \mapsto \min\set{n \mid Y_n(s) = 0}$.}

@@

A main result is the following theorem.
\begin{theorem} 
\label{thm:rankable and strict rankable}
If a closed SPCF term is rankable (respectively, strictly rankable) by $f$, then $(f(M_n))_{n \in \omega}$ is a 
supermartingale (respectively, ranking supermartingale) adapted to the filtration $(\mathcal{F}_n)_{n \in \omega}$ where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
\end{theorem}

\paragraph{Proof of \Cref{thm:rankable and strict rankable}.}
\emph{Notation}. Given an $\omega$-sequence (e.g.~$s \in S$) and $m \in \omega$, we write $s_{\leq m} \in I^m$ to mean the prefix of $s$ of length $m$.

Fix a closed SPCF term $M \in \Lambda^0$, and a ranking function $f$ for it, satisfying $f(N) = 0$ iff $N$ is a value.
Let $n \in \omega$, define the following random variables on the probability space $(S, \calF, \mu)$:
\begin{align*}
M_n(s) &:= \pi_0 (\red^n(M, s))\\
\ndraw{n}{s} &:= |\{k < n \mid \exists E: M_k(s) = E[\tsample]\}|,
\end{align*}
 so that $\pi_1 (\red^n(M, s)) = \overbrace{\pi_t( \cdots (\pi_t}^{\ndraw n s}(s))$, and the filtration $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
The $\calF_n$-measurability of $M_n$ (and hence of $\#\mathrm{draw}_n$) follows from \citep{DBLP:conf/icfp/BorgstromLGS16}.
%$\#\mathrm{draw}_n$ is a stopping time (bounded by $n$). 
\akr{$\#\mathrm{draw}_n$ is not a stopping time. It doesn't look like you actually use this claim anyway, so it should just be fine to remove, but did you mean something different?} \lo{I agree, and I don't actually use this claim.}
Take $s \in A \in \calF_n$ with $\ndraw{n}{s} = l$.
For any $s'\in S$, if $s_{\leq l} = s_{\leq l}'$ then $s' \in A$.
It follows that ${\set{s_{\leq l}} \cdot I^\omega} \subseteq A$.
\lo{NOTE. To fix notation:
\(
\sigma(M_{n}) = \sigma\big(\set{M_{n}^{-1}(\alpha_j, U_{\alpha_j})
\mid \alpha_j\in \mathsf{Sk}_j, U_{\alpha_j} \in \mathcal{B}(\Real^j), j \geq 0}\big)
\)
}
%Take $M \in \Lambda^0$ (closed SPCF terms). Define, for each $n \in \omega$, the random variable $M_n : S \to \Lambda^0$ by $M_n(s) := \pi_0 (\red^n(M, s))$.

@@

\medskip
As before, fix a $M \in \Lambda^0$.
Define random variables on the probability space $(S, \calF, \mu)$:
\begin{align}
T_{-1}(s) & := -1 \nonumber \\
T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value or of form } E[\tY \lambda x. N] \} \nonumber \\
X_n(s) & := f(M_n(s)) \nonumber \\
Y_n(s) & := X_{T_n}(s) \nonumber \\
T_M^\tY(s) &:= \min \set{n \mid M_{T_n(s)}(s) \textrm{ is a value}} \label{eq:y stopping time}
\end{align}
The random variable $T_M^\tY$ is known as the \emph{$\tY$-runtime of $M$}, and it can equivalently be defined as the number of $\tY$-reduction steps in the reduction sequence of $M$. Note that, as the type system ensures that the reduction relation excluding $\tY$-reduction is strongly normalising, only finitely many reductions can occur in a row without one of them being a $\tY$-reduction, therefore $T_M^\tY < \infty$ a.s.~iff $M$ is AST.

\begin{theorem}[Soundness] \label{thm:rankable implies termination}
\begin{enumerate}
\item If a closed SPCF term $M$ is rankable, then $T^{\tY}_M < \infty$ a.s.~(equivalently $M$ is AST), and $\expect{T^{\tY}_M} < \infty$ i.e.~$M$ is $\tY$-positively almost-sure terminating ($\tY$-PAST).

@@

\begin{proof}
For 1, observe that $T_0, T_1, T_2, \cdots$ are an increasing sequence of stopping times adapted to $(\calF_n)_{n \in \omega}$, and each $T_i$ is bounded.\lo{Prove it.}
Thanks to \Cref{thm:rankable and strict rankable}, 
$(X_n)_{n \in \omega}$ is a supermartingale;
it then follows from \Cref{thm:optional sampling} that $(Y_n)_{n \in \omega}$ is a ($1$-ranking) supermartingale with respect to the filtration $(\calF_{T_n})_{n \in \omega}$ and the stopping time $T^{\tY}_M$.
Therefore, by \Cref{lem:rank-PAST}, $T^{\tY}_M < \infty$ a.s.~and $\expect{T^{\tY}_M} < \infty$.

>>>>

\begin{definition}\rm
\begin{enumerate}
\item A sequence of random variables $(Y_n)_{n \in \omega}$ adapted to a filtration $(\calF_n)_{n \in \omega}$ is a \emph{supermartingale} if for all $n \in \omega$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~\changed[lo]{(i.e.~$\forall A \in \calF_n$, $\int_A \dif \mu \, Y_{n+1} \leq \int_A \dif \mu \,Y_n$).}
\item Let $\epsilon > 0$. A stopping time $T$ and a supermartingale $(Y_n)_{n \in \omega}$, both adapted to filtration $(\calF_n)_{n \in \omega}$, are a \emph{$\epsilon$-ranking supermartingale} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.\footnote{For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable $s \mapsto \begin{cases}
1 & \hbox{if $s \in A$}\\
0 & \hbox{otherwise}
\end{cases}$.}
A \emph{ranking supermartingale} is a $\epsilon$-ranking supermartingale for some $\epsilon > 0$.
\end{enumerate}
\end{definition}

\begin{remark}
\changed[lo]{The original definition of ranking supermartingale, due to \citep{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, does not refer to a stopping time explicitly:
Fioriti and Hermanns call a supermartingale $(Y_n)_{n \in \omega}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot 1_{\set{Y_n > 0}}$.}
\end{remark}

@@

The following is an iterated version of Doob's well-known Optional Sampling Theorem (see, e.g., \cite[\S 6.7]{AshDD00} and \cite[Theorem 7.2]{DBLP:conf/popl/FioritiH15}).
\begin{theorem}[Optional Sampling]
\label{thm:optional sampling}
Let $(X_n)_{n\in \omega}$ be a supermartingale, and $(T_n)_{n \in \omega}$ a sequence of increasing stopping times, both adapted to filtration $(\calF_n)_{n \in \omega}$, then $(X_{T_n})_{n \in \omega}$ is a supermartingale adapted to $(\calF_{T_n})_{n \in \omega}$ if one of the following conditions holds:
\begin{enumerate}
\item each $T_n$ is bounded i.e.~$T_n < c_n$ where $c_n$ is a constant
\item $(X_n)_{n\in \omega}$ is uniformly integrable.
\end{enumerate}
\end{theorem}

\begin{lemma}%[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}}]
% \citep[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}
\label{lem:rank-PAST}
Let $(Y_n)_{n \in \omega}$ be a $\epsilon$-ranking supermartingale with respect to the stopping time $T$, then $T < \infty$ almost surely, and $\expect{T} \leq \dfrac{\expect{Y_0}}{\epsilon}$.
\end{lemma}

\begin{proof}
We first prove by induction on $n$: 
\[
\forall n \in \omega \, . \, \expect{Y_n} \leq \expect{Y_0} - \epsilon \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[T > i]\Big).
\]
It follows that $\sum_{i=0}^{\infty} \mathbb{P}[T > i]$ converges; hence $\lim_{i \to \infty} \mathbb{P}[T > i] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
It then remains to observe: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=0} \mathbb{P}[T > i]$.
\end{proof}

\changed[lo]{\Cref{lem:rank-PAST} is a slight extension of \citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}. 
In Lemma 5.5 of op.~cit., the stopping time $T$ is assumed to be the random variable $s \mapsto \min\set{n \mid Y_n(s) = 0}$.}

@@

Our first result is the following theorem.
\begin{theorem}[Rankability] 
\label{thm:rankable and strict rankable}
If a closed SPCF term is rankable (respectively, strictly rankable) by $f$, then $(f(M_n))_{n \in \omega}$ is a 
supermartingale (respectively, ranking supermartingale) adapted to the filtration $(\mathcal{F}_n)_{n \in \omega}$ where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.
\end{theorem}

%\subsection{Proof of \Cref{thm:rankable and strict rankable}.}
\emph{Notation}. Given an $\omega$-sequence (e.g.~$s \in S$) and $m \in \omega$, we write $s_{\leq m} \in I^m$ to mean the prefix of $s$ of length $m$.

Fix a closed SPCF term $M \in \Lambda^0$, and a ranking function $f$ for it, satisfying $f(N) = 0$ iff $N$ is a value.
Let $n \in \omega$, define the following random variables on the probability space $(S, \calF, \mu)$:
\begin{align*}
M_n(s) &:= \pi_0 (\red^n(M, s))\\
\ndraw{n}{s} &:= |\{k < n \mid \exists E: M_k(s) = E[\tsample]\}|,
\end{align*}
 so that $\pi_1 (\red^n(M, s)) = \overbrace{\pi_t( \cdots (\pi_t}^{\ndraw n s}(s))$, and the filtration $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
The $\calF_n$-measurability of $M_n$ (and hence of $\#\mathrm{draw}_n$) follows from \citep{DBLP:conf/icfp/BorgstromLGS16}.
%$\#\mathrm{draw}_n$ is a stopping time (bounded by $n$). 
\iffalse
\akr{$\#\mathrm{draw}_n$ is not a stopping time. It doesn't look like you actually use this claim anyway, so it should just be fine to remove, but did you mean something different?} \lo{I agree, and I don't actually use this claim.}
\fi
Take $s \in A \in \calF_n$ with $\ndraw{n}{s} = l$.
For any $s'\in S$, if $s_{\leq l} = s_{\leq l}'$ then $s' \in A$.
It follows that ${\set{s_{\leq l}} \cdot I^\omega} \subseteq A$.
\lo{NOTE. To fix notation:
\(
\sigma(M_{n}) = \sigma\big(\set{M_{n}^{-1}(\alpha_j, U_{\alpha_j})
\mid \alpha_j\in \mathsf{Sk}_j, U_{\alpha_j} \in \mathcal{B}(\Real^j), j \geq 0}\big)
\)
}
%Take $M \in \Lambda^0$ (closed SPCF terms). Define, for each $n \in \omega$, the random variable $M_n : S \to \Lambda^0$ by $M_n(s) := \pi_0 (\red^n(M, s))$.

@@

\medskip
As before, fix a closed SPCF term $M \in \Lambda^0$.
Define random variables on the probability space $(S, \calF, \mu)$:
\begin{align}
T_{-1}(s) & := -1 \nonumber \\
T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value or of form } E[\tY \lambda x. N] \} \nonumber \\
X_n(s) & := f(M_n(s)) \nonumber \\
Y_n(s) & := X_{T_n}(s) \nonumber \\
T_M^\tY(s) &:= \min \set{n \mid M_{T_n(s)}(s) \textrm{ is a value}} \label{eq:y stopping time}
\end{align}
The random variable $T_M^\tY$, which we call the \emph{$\tY$-runtime of $M$}, can equivalently be defined as the number of $\tY$-reduction steps in the reduction sequence of $M$. 
Note that, as the type system ensures that the reduction relation excluding $\tY$-reduction is strongly normalising, only finitely many reductions can occur in a row without one of them being a $\tY$-reduction, therefore $T_M^\tY < \infty$ a.s.~iff $M$ is AST.

The r.v.~$T_M^\tY$ is a stopping time adapted to the filtration $(\calF_{T_n})_{n \in \omega}$.

\begin{theorem}[Soundness] \label{thm:rankable implies termination}
\begin{enumerate}
\item If a closed SPCF term $M$ is rankable, then $M$ is AST (equivalently, $T^{\tY}_M < \infty$ almost surely), and $M$ is $\tY$-positively almost-sure terminating ($\tY$-PAST) i.e.~$\expect{T^{\tY}_M} < \infty$.

@@

\begin{proof}
Let $f$ be a ranking function on $M$.
For 1, observe that $T_0, T_1, T_2, \cdots$ are an increasing sequence of stopping times, each is bounded \lo{Prove it.} and adapted to $(\calF_n)_{n \in \omega}$.
Then, since $(f(M_n))_{n \in \omega}$ is a supermartingale also adapted to $(\calF_n)_{n \in \omega}$ (\Cref{thm:rankable and strict rankable}),
it follows from the Optional Sampling \Cref{thm:optional sampling} that $(Y_n)_{n \in \omega}$ where $Y_n := f(M_{T_n})$ is a supermartingale adapted to the filtration $(\calF_{T_n})_{n \in \omega}$.
Since $T_M^\tY$ is a stopping time also adapted to $(\calF_{T_n})_{n \in \omega}$, we have that $(Y_n)_{n \in \omega}$ is a $1$-ranking supermartingale.
Therefore, by \Cref{lem:rank-PAST}, $T^{\tY}_M < \infty$ a.s.~and $\expect{T^{\tY}_M} < \infty$.
